Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-05 code/bias_niche.Rmd
update

## 1.3 Evaluate models
```{r}
# Summary
summary(brm_pvis00)
summary(brm_pvis01)
summary(brm_pvis02)
summary(brm_pvis11)
summary(brm_pvis21)

# Posterior predictive checks
pp_check(brm_pvis00)
pp_check(brm_pvis01)
pp_check(brm_pvis02)
pp_check(brm_pvis11)
pp_check(brm_pvis21)

# Bayes R2
bayes_R2(brm_pvis00)
bayes_R2(brm_pvis01)
bayes_R2(brm_pvis02)
bayes_R2(brm_pvis11)
bayes_R2(brm_pvis21)

@@

ggplot(brm_pvis01_plotframe1, aes(condition_mean_trans, ktype, 
                                 xmin = .lower_trans, xmax = .upper_trans)) +
  geom_pointinterval() +
  theme_light(14) +
  labs(x = "Visitation probability", y= "Morphotype") +
  guides(fill = FALSE)

ggsave(file = "../output/brm_pvis01_plot1.pdf", width = 5, height = 5)

@@

ggplot(bb_distribution, aes(reorder(bb.sp, elev.med), elev.mean)) +
  geom_line(size = 2, alpha = 0.25) +
  geom_point(aes(size = abund), alpha = 0.5) +
  facet_wrap(~year, ncol = 1) +
  theme_light(12) +
  scale_size_continuous(name = "Abundance") +
  xlab("BB species") +
  ylab("Elevation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

get_variables(brm_niche00)
conditional_effects(brm_niche00)
mcmc_plot(brm_niche00, pars = c("^r_sp.pair\\[mend*"))
mcmc_plot(brm_niche00, pars = c("^r_site*"))
```

>>>>

## 1.3 Evaluate models
```{r}
# Summary
summary(brm_pvis00)
summary(brm_pvis01)
summary(brm_pvis02)

# Posterior predictive checks
pp_check(brm_pvis00)
pp_check(brm_pvis01)
pp_check(brm_pvis02)

# Bayes R2
bayes_R2(brm_pvis00)
bayes_R2(brm_pvis01)
bayes_R2(brm_pvis02)
```

@@

ggplot(brm_pvis01_plotframe1, aes(condition_mean_trans, ktype, 
                                 xmin = .lower_trans, xmax = .upper_trans)) +
  geom_pointinterval() +
  theme_light(16) +
  labs(x = "Visitation probability", y= "Morphotype") +
  guides(fill = FALSE)

ggsave(file = "../output/brm_pvis01_plot1.pdf", width = 6, height = 7)

@@

ggplot(bb_distribution, aes(reorder(bb.sp, elev.med), elev.mean)) +
  geom_line(size = 2, alpha = 0.25) +
  geom_point(aes(size = abund), alpha = 0.5) +
  facet_wrap(~year, ncol = 1) +
  theme_light(12) +
  scale_size_continuous(name = "Abundance") +
  xlab("BB species") +
  ylab("Elevation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

bb_distribution_pool <- visitation %>%
  left_join(site_data) %>%
  group_by(site, date, elev.mean, bb.sp) %>%
  summarize(abund = n()) %>%
  ungroup() %>%
  complete(site, date, elev.mean, bb.sp, fill = list(abund = 0)) %>%
  group_by(site, elev.mean, bb.sp) %>%
  summarize(abund = mean(abund)) %>%
  filter(abund > 0) %>%
  group_by(bb.sp) %>%
  mutate(elev.floor = min(elev.mean),
         elev.ceiling = max(elev.mean),
         elev.range = elev.ceiling - elev.floor,
         elev.med = elev.floor + ((elev.ceiling - elev.floor)/2))

ggplot(bb_distribution_pool, aes(reorder(bb.sp, elev.med), elev.mean)) +
  geom_line(size = 2, alpha = 0.25) +
  geom_point(aes(size = abund), alpha = 0.5) +
  theme_light(14) +
  scale_size_continuous(name = "Abundance") +
  xlab("BB species") +
  ylab("Elevation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../output/bb_range.pdf", width = 6, height = 5)
```

@@

get_variables(brm_niche00)
conditional_effects(brm_niche00)
mcmc_plot(brm_niche00, pars = c("^r_sp.pair\\[mend*"))
mcmc_plot(brm_niche00, pars = c("^r_site*"))

brm_niche00_plotframe1 <- brm_niche00 %>%
  spread_draws(b_Intercept, b_levelgenus, b_levelmorphotype) %>%
  mutate(b_levelgenus = b_Intercept + b_levelgenus,
         b_levelmorphotype = b_Intercept + b_levelmorphotype) %>%
  pivot_longer(-c(.chain, .iteration, .draw), names_to = "level", values_to = "value") %>%
  group_by(level) %>%
  point_interval(value, .width = c(0.95, 0.66)) %>% # calculate credible intervals
  mutate(value_trans = inv_logit_scaled(value), # exponentiating the estimates gets us back to the response scale (negbinomial uses log link function)
         .lower_trans = inv_logit_scaled(.lower),
         .upper_trans = inv_logit_scaled(.upper)) %>%
  mutate(level = case_when(
    level == "b_Intercept" ~ "family",
    level == "b_levelmorphotype" ~ "morphotype",
    level == "b_levelgenus" ~ "genus"
  )) %>%
  mutate(level = factor(level, levels = c("genus", "family", "morphotype")))

ggplot(brm_niche00_plotframe1, aes(value_trans, level, 
                                 xmin = .lower_trans, xmax = .upper_trans)) +
  geom_pointinterval() +
  theme_light(16) +
  labs(x = "Niche overlap", y= "Level") +
  guides(fill = FALSE)

ggplot(brm_niche00_plotframe1, aes(level, value_trans, 
                                 ymin = .lower_trans, ymax = .upper_trans)) +
  geom_pointinterval() +
  theme_light(16) +
  labs(x = "Level", y= "Niche overlap") +
  guides(fill = FALSE) +
  ylim(c(0, 0.55))

#ggsave(file = "../output/brm_pvis01_plot1.pdf", width = 6, height = 7)
```

### 2021-03-05 ms_2/ms_2_v2.Rmd
update

The role of competition in regulating community assembly and species coexistence has long been a central question in ecology. It has taken on new urgency, however, on account of recent climate change, which threatens to disrupt existing patterns of species interactions [@Tylianakis2008-fi], including competition [@Ettinger2017-pk], with consequences for community composition and function. These effects are likely to be most pronounced along the steep climate gradients of mountain slopes, where climate-induced range shifts are expected to introduce novel competitive interactions within trophic levels and cause spatial or phenological mismatches between organisms and their food resources.

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, bumble bees have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phylogeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. Teasing apart the still-obscure mechanisms of bumble bee coexistence and niche partitioning remains a salient problem in basic ecology, but it has also become an urgent question in the context of conservation. Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. Warming climates have already driven measurable elevational range shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and these are predicted to cause declines or even extinction in many bumble bees species [@Soroye2020-zb]. One mechanism by which climate change is expected to affect bumble bees is via the introduction of novel interspecific competition driven by the advance of lowland bumble bee species into the elevational distributions of their highland counterparts. Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales.

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in fundamental dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the morphological fittedness between bumble bee species and their floral hosts. Bumble bees and flowers vary in an array of interrelated morphological parameters, both within and between species, but for bumble bees, morphological variation consists principally (though not exclusively) of variation in tongue length. Floral handling time depends on the degree of match between bumble bee tongue length and floral corolla depth [@Inouye1980-mv], and there is a well-established (though imperfect) association between tongue length and floral choice. Importantly, though, accessing floral reward involves more than just a match between tongue length and corolla depth, and in some flowers access to floral rewards involves learning to manipulate floral structures that are challenging irrespective of tongue length (Laverty 1994). For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries (Westerkamp 1997), and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees (Wallace 1977). Thus the dietary niche of a given bumble bee species in a given ecological context is an emergent product of innate floral biases, rewards experienced during foraging, and learned flower handling skills, and all of which contribute to the phenomenon of interspecific niche overlap upon which competition is predicated.

Empirical measurements of niche overlap, however, are inherently equivocal in their interpretation with respect to competition [@Pianka1972-oi; @Pianka1974-qk; @Abrams1980-am]. Low niche overlap between two species, for example, could be interpreted prospectively as an indication that the *potential* for interspecific competition is low, but the same pattern could also be interpreted retrospectively as a mitigating response to high interspecific competition. Which of these interpretations is most plausible hinges on how closely species' observed niches approximate their fundamental niches (within a given resource context). Where observed niches are close to absolute niches, low niche overlap is an indication of low potential for interspecific competition. Any difference between observed and fundamental niches, though, might be attributable to competition-driven niche partitioning, and niche overlap as a reflection of potential competition cannot be distinguished from niche overlap as a response to realized competition. In other words, niche overlap can only be intepreted as an index of *potential* competition if *realized* competition is assumed to be negligible, and this assumption can only be justified on the basis of additional knowledge of the study system under consideration.

Rather than considering various floral morphological traits independently, we rely on the synthesis of Kugler (1970), who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice (Kugler 1943), and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes (Neumayer and Paulus 1999; Schneller et al. 2014).

<!-- 
That one paper [@Zhongming2020-xj] simply equated diet overlap with competition, and I think that is a mistake. But I think we can make a more nuanced case for its use as an index of potential competition. Indeed, Ingolf interpreted diet overlap in exactly this sense in one of his early papers [@Steffan-Dewenter2000-ll]. In fact, there is a slew of papers, from across a wide range of study systems and organisms, that use the term "potential competition" to interpret diet overlap. I think I'm on solid ground.
-->

Importantly, our study system is one of rich floral resources and harsh abiotic conditions, and it is likely that competition is low, or at least transient, and that bumble bees are limited instead by abiotic conditions during the critical period of spring colony founding. Thus, the patterns in our data likely represent floral choice unconstrained by competition and therefore approximate the absolute dietary niches of the species involved (within the context of the studied floral community). On this basis, we argue for the interpretation of niche overlap as an index of *potential* competition rather than a response to *realized* competition, with the caveat this interpretation may misrepresent certain temporal intervals (e.g. early spring colony founding) or elevational bands (e.g. the sparse alpine grasslands above the tree line ecotone, which in our system falls around 1500 m).

We first ask how floral biases compare across bumble bee species whether these patterns can be explained by implicit trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of dietary niche overlap between bumble bee species that emerge from their respective floral biases, comparing niche overlap at the nested levels of floral morphotype, floral family, and floral genus. Finally, we ask how niche overlap varies with elevation. With respect to elevation, we analyze both pairwise niche overlap and community mean niche overlap, since low pairwise niche overlap could conceal a high potential for diffuse competition [@Pianka1974-qk; MacArthur1972-qz]. We discuss the implications of these findings for bumble bee conservation in a changing climate

@@

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures in a coniferous forest matrix. We selected 25 study sites (60 x 60 m) on pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management. The sites sampled in this study are the same as those used in a series of previous studies [@Hoiss2012-sn; @Hoiss2012-kr; @Hoiss2015-gv], but the data reported here are otherwise independent.

@@

@@

Floral morphology was summarized using floral morphotypes of Kugler (1970), accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, so this morphotype was also omitted from all analyses.

@@

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years, for the reasons described above, make comparisons across years difficult to interpret, and we deemed these comparisons not worth the cost of added model complexity.

<!-- Should I be including year as a varying effect in these models? -->

@@

<!-- Consider adding year as a varying effect -->

@@

<!-- Year should really be included in Model 5, or just use 2012. -->

@@

@@

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species groups lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

@@

```{r fig1, echo = FALSE,  fig.cap = paste("Each floral morphotype is exemplified by a representative species common in our study system. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.")}
knitr::include_graphics("../output/ktype_abund_gam_mod.png")
```

@@

```{r fig2, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by floral morphotype. Point-interval plots depict 66\\% and 95\\% CIs.")}
knitr::include_graphics("../output/brm_pvis01_plot1.pdf")
```

@@

Disc (group 1) flowers were visited at similar rates across tongue length classes. At the species level, disc flowers were visited at relatively high rates by the short-tongued B. terrestris/lucorum and the medium-tongued Bombus monticola. Visitation to funnel (group 2) flowers was dominated by long-tongued species, an effect driven entirely by the Aconitum (group 2) specialist B. gerstaeckeri. Bell (group 3) flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were visited at highest probability by the short-tongued species B. jonellus, B. pyrenaeus, and B. soroeensis, and by the medium-tongued species B. hypnorum and B. monticola. Stalk-disc (group 4) flowers were visited at similar rates across all tongue length classes, and there was also broad overlap in visitation rates at the species level. Lip (group 5) flowers were also visited at similar rates across tongue length classes, but at the species level B. pascuorum exhibited an uniquely high preference for lip flowers. Flag (group 6) flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued B. pascuorum and B. mucidus, moderately high visitation from the short-tongued B. terrestris/lucorum and B. lapidarius, the medium-tongued B. monticola, and the long-tongued B. mendax and B. hortorum, and low visitation by the short-tongued B. soroeensis, B. pyrenaeus, B. psithyrus, B. pratorum, and B. jonellus, the medium-tongued B. hypnorum, and the long-tongued B. gerstaeckeri.

@@

## The variation of morphological bias through elevation and time

Model 5 explained 33.8% of total variance (adjusted R2 = 0.37). All tensor product smooths (Figure 5) were significant at p < 0.001 except for the one representing B. pratorum and flag (group 6), which was more weakly significant at p = 0.02. As in Models 1-3, the results of Model 5 should be interpreted in terms of conditional visitation probability.

Bell (group 3) flowers were visited at highest rates early in the season and at low elevation by all species except B. hortorum, whose visitation rate to bell flowers peaked in mid-season and at mid-elevation. Notably, the medium-tongued B. pascuorum and B. wurflenii also exhibited relatively high visitation to bell flowers at mid-season and mid-elevation, while the visitation rates of the short-tongued B. pratorum, B. soroeensis. and B. terrestris lucorum were more concentrated at the early-season and low-elevation peak.

Lip (group 5) flowers exhibited roughly the mirror image of the pattern of visitation rate seen in bell flowers, with peak visitation late in the season and at low elevation, but relatively high visitation extending toward mid-season and mid-elevation. This pattern was consistent for all bumble bee species.

The pattern of visitation rates to flag (group 6) flowers fell generally along the lines of tongue length class. The short-tongued B. pratorum and B. soroeensis exhibited peak visitation early in the season and at low elevation. The long-tongued B. hortorum and the medium-tongued B. pascuorum and B. wurflenii exhibited peak visitation at mid-season, though B. pascuorum exhibited a bimodal pattern in elevation, with peaks around 1000 m and 1750 m, while the visitation rate of B. hortorum declined monotonically from high elevation to low. The short-tongued B. terrestis/lucorum broke the tongue-length pattern, exhibiting a diffuse pattern of visitation rate more similar to that of the long- and medium-tongued species than the other short-tongued species.

Visitation rates for head (group 7) flowers were more or less bimodal for all species, with a first peak in mid-season and low elevation and a second peak in late-season and high elevation.

```{r fig6, echo = FALSE,  fig.cap = paste("GAM analysis of bias as a function of the tensor-product interaction of elevation and day of year, parsed by bumble bee species and floral morphotype and conditioned on the abundance of each bumble bee species and the relative abundance of each floral morphotype. The color ramp of each smooth has been scaled to equal color range, thereby obscuring the absolute magnitude of effect. To avoid cluttering the figure, we have opted not to include a unique color ramp key for each plot, since the focus of our inference is on relative rather than absolute patterns.")}
knitr::include_graphics("../output/ktype_lability_gam_mod.png")
```

@@

For bumble bees, floral bias is a labile product of innate preferences, the soft constraints of morphological compatibility, and the dynamic ecological context in which foraging occurs. By conditioning bumble bee visitation data upon concurrent floral abundance data across spatially and temporally replicated samples, we demonstrate the strength bumble bee floral bias, its functional underpinnings, its variation across species, and its lability within species in response to environmental variation. Such tuning of floral bias has obvious adaptive significance for bumble bees, which are characteristically — and most likely primordially — associated with mountain habitats (Williams 1985; Hines 2008), where steep temperature gradients and short growing seasons generate extreme spatial and temporal floral turnover.

>>>>

The role of competition in regulating community assembly and species coexistence has long been a central question in ecology. It has taken on new urgency, however, on account of recent climate change, which threatens to disrupt existing patterns of species interactions [@Tylianakis2008-fi], including competition [@Ettinger2017-pk], with consequences for community composition and function. These effects are likely to be most pronounced along the steep climate gradients of mountain slopes, where rapid climate-induced range shifts are expected to introduce novel competitive interactions within trophic levels and cause spatial or phenological mismatches between organisms and their food resources.

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phylogeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. Teasing apart the still-obscure mechanisms of bumble bee coexistence and niche partitioning remains a salient problem from a basic ecological perspective, but it has also become an urgent question in the context of applied conservation. Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. Warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and these are predicted to cause declines or even extinction in many bumble bees species [@Soroye2020-zb]. One mechanism by which climate change is expected to affect bumble bees is the introduction of novel interspecific competition driven by the upslope advance of lowland bumble bee species into the elevational distributions of their highland counterparts. Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales and instrumental in the future viability of bumble bee species and their floral mutualists.

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the morphological fittedness between bumble bees and their floral hosts, which consists principally in the compatibility between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases both for short-tongued bumble bees foraging on deep flowers and long-tongued bees foraging on shallow flowers [@Inouye1980-mv], which leads to an association between tongue length and floral niche in bumble bee communities [e.g. @Inouye1978-qi; @Inouye1980-mv]. This association between tongue length and floral morphology, though, is imperfect [e.g. @Ranta1980-ok; @Goulson2008-oe], and access to floral rewards in some flowers involves learning to manipulate floral structures that are challenging irrespective of tongue length [@Laverty1994-hm]. For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries [@Westerkamp1997-hs], and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees [@Wallace1977-kz]. Thus the dietary niches of bumble bee species in a given ecological context are emergent products of innate floral biases, rewards experienced during foraging, and learned flower handling skills, all of which contribute to the phenomenon of interspecific niche overlap upon which interspecific competition is predicated.

<!-- I expect my co-authors to want the section below to get pushed to the discussion. I disagree, but I think we could make it work if they insist. -->

Empirical measurements of niche overlap are, however, inherently ambiguous with respect to competition. It has been well-demonstrated that niche overlap can never serve as a metric of competition *per se* [@Abrams1980-am], but it is often used as an index of *potential* competition, particularly in the context of dietary studies [e.g. @French1985-eo; @Steffan-Dewenter2000-ll; @Jung2015-cw], based on the reasoning that if and when food becomes limiting, competition can be expected to vary in proportion to the dietary overlap of co-occurring species [@Pianka1972-oi]. Focusing on *potential* competition does not, however, escape the equivocity of niche overlap, since an observed pattern of niche overlap could just as plausibly be interpreted as adaptive niche partitioning in response to *realized* competition [@Pianka1974-qk]. This creates a predicament in which niche overlap can only be interpreted as an index of potential competition when realized competition --- which cannot be inferred from niche overlap --- is assumed to be negligible, and this assumption can only be justified on the basis of additional knowledge of the study system under consideration. Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows --- generated by past or ongoing grazing management --- embedded in a matrix of coniferous forest. This system is one of rich floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bees are limited instead by abiotic conditions during the critical period of spring colony founding. On this basis, we argue for the interpretation of niche overlap as an index of *potential* competition rather than a response to *realized* competition, with the caveats this interpretation may misrepresent certain temporal intervals (e.g. early spring colony founding) or elevational bands (e.g. the sparse alpine grasslands above the tree line ecotone) [Sponsler et al., submitted] and that the residual effects of past competitive pressure on innate floral biases [@Connell1980-dn] are beyond the scope of our inference.

The aim of our study is to synthesize the phenomena of floral bias, dietary niche overlap, and interspecific competition in a way that accounts for the coexistence of bumble bee species with similar tongue lengths and predicts the effects of climate-driven elevational range shifts on the future of potential competition between bumble bee species in mountain habitats. We first ask how floral biases compare across bumble bee species and to what extent these patterns can be explained by trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of dietary niche overlap between bumble bee species that emerge from their respective floral biases, comparing niche overlap at the nested levels of floral morphotype, floral family, and floral genus. Finally, we ask how niche overlap varies with elevation and discuss the implications of this variation for bumble bee conservation in a changing climate.

@@

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures in a coniferous forest matrix. We selected 25 study sites (60 x 60 m) on pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management.

@@

## Data sources

### Floral morphology

Rather than considering various floral morphological traits independently, we rely on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Neumayer1999-il; @Schneller2014-ry].

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, so this morphotype was also omitted from all analyses.

### Bumble bee tongue length

@@

@@

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years, for the reasons described above, would make comparisons across years difficult to interpret, and we deemed these comparisons not worth the cost of added model complexity. This decision entails the caveats that our analysis cannot detect year-to-year variation in floral biases and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value.

@@

@@

@@

### Defining bias

We opt to use of the term “bias” in favor of the term “preference,” since the latter is laden with more assumptions and its inference under field conditions is problematic [@Williams2005-yp]. By “bias,” we mean simply an empirical disproportionality between the relative abundance of a given floral taxon or group and its proportional visitation. The exact mechanism of this disproportionality — whether an arbitrary “preference” on the part of the foraging bee, an unmeasured decoupling of relative abundance and rate of reward, etc. — is beyond the scope of our discussion, and probably best studied under laboratory conditions where strict factorial control is possible. We focus instead on the empirical phenomenon of bias, its variation, and its correlation to morphology.

@@

We analyze both pairwise niche overlap and community mean niche overlap, since low pairwise niche overlap could conceal a high potential for diffuse competition [@Pianka1974-qk; MacArthur1972-qz]

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species groups lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

@@

@@

```{r fig1, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by floral morphotype. Point-interval plots depict 66\\% and 95\\% CIs. Each floral morphotype is exemplified by a representative species common in our study system. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.")}
knitr::include_graphics("../output/brm_pvis01_plot1_mod.png")
```

@@

Disc (group 1) flowers were visited at similar rates across tongue length classes. At the species level, disc flowers were visited at relatively high rates by the short-tongued B. terrestris/lucorum and the medium-tongued Bombus monticola. Visitation to funnel (group 2) flowers was dominated by long-tongued species, an effect driven entirely by the Aconitum (group 2) specialist B. gerstaeckeri. Bell (group 3) flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were visited at highest probability by the short-tongued species B. jonellus, B. pyrenaeus, and B. soroeensis, and by the medium-tongued species B. hypnorum and B. monticola. Stalk-disc (group 4) flowers were visited at similar rates across all tongue length classes, and there was also broad overlap in visitation rates at the species level. Lip (group 5) flowers were also visited at similar rates across tongue length classes, but at the species level B. pascuorum exhibited an uniquely high bias toward lip flowers. Flag (group 6) flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued B. pascuorum and B. mucidus, moderately high visitation from the short-tongued B. terrestris/lucorum and B. lapidarius, the medium-tongued B. monticola, and the long-tongued B. mendax and B. hortorum, and low visitation by the short-tongued B. soroeensis, B. pyrenaeus, B. psithyrus, B. pratorum, and B. jonellus, the medium-tongued B. hypnorum, and the long-tongued B. gerstaeckeri.

@@

@@

For bumble bees, floral bias is a labile product of innate biases, the soft constraints of morphological compatibility, and the dynamic ecological context in which foraging occurs. By conditioning bumble bee visitation data upon concurrent floral abundance data across spatially and temporally replicated samples, we demonstrate the strength bumble bee floral bias, its functional underpinnings, its variation across species, and its lability within species in response to environmental variation. Such tuning of floral bias has obvious adaptive significance for bumble bees, which are characteristically — and most likely primordially — associated with mountain habitats (Williams 1985; Hines 2008), where steep temperature gradients and short growing seasons generate extreme spatial and temporal floral turnover.
